\documentclass[11pt]{article}
\usepackage{amsmath,amssymb}
\usepackage{hyperref}
\usepackage[margin=0.7in]{geometry}
\usepackage{booktabs}
\usepackage{graphicx}

\title{Assignment 4: CS 754, Advanced Image Processing\\
\large Solutions to Questions 1, 4, and 5}
\author{}
\date{}

\begin{document}
\maketitle

%======================================================================
\section*{Question 1: Proof of the Eckart-Young Theorem (Exercise 7.2)}
%======================================================================

We prove the Eckart-Young theorem, which states that the best rank-$k$
approximation to a matrix in the Frobenius norm is given by its truncated SVD.

\subsection*{1.1 Theorem Statement}

\textbf{Theorem (Eckart-Young, 1936).}
Let $\boldsymbol{A} \in \mathbb{R}^{m \times n}$ with $\operatorname{rank}(\boldsymbol{A}) = r$
and singular value decomposition
\[
\boldsymbol{A} = \boldsymbol{U}\boldsymbol{\Sigma}\boldsymbol{V}^T
= \sum_{i=1}^{r} \sigma_i \,\boldsymbol{u}_i \boldsymbol{v}_i^T,
\]
where $\sigma_1 \geq \sigma_2 \geq \cdots \geq \sigma_r > 0$.
For any $k \leq r$, define
\[
\boldsymbol{A}_k = \sum_{i=1}^{k} \sigma_i \,\boldsymbol{u}_i \boldsymbol{v}_i^T
= \boldsymbol{U}_k \boldsymbol{\Sigma}_k \boldsymbol{V}_k^T,
\]
where $\boldsymbol{U}_k$, $\boldsymbol{\Sigma}_k$, $\boldsymbol{V}_k$ contain only
the first $k$ columns/entries. Then $\boldsymbol{A}_k$ solves
\[
\min_{\boldsymbol{B}:\;\operatorname{rank}(\boldsymbol{B})\leq k}
\|\boldsymbol{A} - \boldsymbol{B}\|_F^2,
\]
and the minimum value is $\sum_{i=k+1}^{r}\sigma_i^2$.

\subsection*{1.2 Proof}

The proof uses the \emph{unitary invariance} of the Frobenius norm:
for any orthogonal $\boldsymbol{P} \in \mathbb{R}^{m \times m}$ and
$\boldsymbol{Q} \in \mathbb{R}^{n \times n}$,
\[
\|\boldsymbol{P}\boldsymbol{M}\boldsymbol{Q}\|_F = \|\boldsymbol{M}\|_F.
\]

\paragraph{Step 1: Reduction via unitary invariance.}
Let $\boldsymbol{B}$ be any $m \times n$ matrix with $\operatorname{rank}(\boldsymbol{B}) \leq k$.
Since $\boldsymbol{A} = \boldsymbol{U}\boldsymbol{\Sigma}\boldsymbol{V}^T$, we write
\[
\|\boldsymbol{A} - \boldsymbol{B}\|_F
= \|\boldsymbol{U}\boldsymbol{\Sigma}\boldsymbol{V}^T - \boldsymbol{B}\|_F
= \|\boldsymbol{U}^T(\boldsymbol{U}\boldsymbol{\Sigma}\boldsymbol{V}^T - \boldsymbol{B})\boldsymbol{V}\|_F
= \|\boldsymbol{\Sigma} - \boldsymbol{U}^T\boldsymbol{B}\boldsymbol{V}\|_F.
\]
Define $\boldsymbol{N} := \boldsymbol{U}^T \boldsymbol{B}\boldsymbol{V}$.
Since $\boldsymbol{U}$ and $\boldsymbol{V}$ are orthogonal, $\operatorname{rank}(\boldsymbol{N}) = \operatorname{rank}(\boldsymbol{B}) \leq k$.
The problem reduces to
\[
\min_{\boldsymbol{N}:\;\operatorname{rank}(\boldsymbol{N})\leq k}
\|\boldsymbol{\Sigma} - \boldsymbol{N}\|_F^2.
\]

\paragraph{Step 2: Expand the Frobenius norm.}
Since $\boldsymbol{\Sigma}$ is an $m \times n$ matrix with $\Sigma_{ii} = \sigma_i$ for $i = 1,\dots,r$ and all other entries zero, we have
\begin{align}
\|\boldsymbol{\Sigma} - \boldsymbol{N}\|_F^2
&= \sum_{i=1}^{r}(\sigma_i - N_{ii})^2
 + \sum_{i=r+1}^{\min(m,n)} N_{ii}^2
 + \sum_{i \neq j} N_{ij}^2. \label{eq:expand}
\end{align}
Each of the three terms is a sum of squared real numbers and hence non-negative.

\paragraph{Step 3: Lower bound via diagonal restriction.}
Any $m \times n$ matrix $\boldsymbol{N}$ of rank at most $k$ has at most $k$ non-zero rows
(or more precisely, its diagonal can have at most $k$ non-zero entries if we want to simultaneously
zero out all off-diagonal terms). Observe:
\begin{itemize}
    \item The third term $\sum_{i\neq j}N_{ij}^2 \geq 0$ is minimised when $N_{ij}=0$ for all $i\neq j$
          (i.e., $\boldsymbol{N}$ is diagonal).
    \item The second term $\sum_{i=r+1}^{\min(m,n)} N_{ii}^2 \geq 0$ is minimised when $N_{ii}=0$ for $i > r$.
    \item Among diagonal $\boldsymbol{N}$ with rank $\leq k$, at most $k$ diagonal entries can be non-zero.
\end{itemize}
So the problem reduces to choosing at most $k$ of the indices $\{1,\dots,r\}$ at which to
set $N_{ii} = \sigma_i$, with $N_{ii} = 0$ at all remaining indices.

\paragraph{Step 4: Optimal index selection.}
Setting the off-diagonal terms and the terms for $i > r$ to zero, we need to minimise
\[
\sum_{i=1}^{r}(\sigma_i - N_{ii})^2
\quad\text{subject to at most $k$ of the $N_{ii}$ being nonzero and each }\leq \sigma_i.
\]
For a given index $i$ where we \emph{do} set $N_{ii} \neq 0$, the optimal choice is $N_{ii} = \sigma_i$ (reducing that squared term to zero).
For indices where $N_{ii} = 0$, the contribution is $\sigma_i^2$.
To minimise the total, we should zero out the $k$ \emph{largest} squared terms, i.e., set $N_{ii} = \sigma_i$ for $i = 1, 2, \dots, k$ (since $\sigma_1 \geq \sigma_2 \geq \cdots \geq \sigma_r$).

\paragraph{Step 5: Conclude.}
The optimal $\boldsymbol{N}^*$ is the diagonal matrix with entries
$N^*_{ii} = \sigma_i$ for $i=1,\dots,k$ and $N^*_{ii}=0$ otherwise.
Mapping back: $\boldsymbol{B}^* = \boldsymbol{U}\boldsymbol{N}^*\boldsymbol{V}^T = \sum_{i=1}^k \sigma_i\,\boldsymbol{u}_i\boldsymbol{v}_i^T = \boldsymbol{A}_k$.

The minimum approximation error is
\[
\|\boldsymbol{A} - \boldsymbol{A}_k\|_F^2
= \sum_{i=k+1}^{r}\sigma_i^2.
\]

\hfill$\blacksquare$

\subsection*{1.3 Remarks}

\begin{enumerate}
\item The elegance of this proof lies in the fact that unitary invariance of the Frobenius norm
completely decouples the optimisation into independent scalar problems on each entry
of the rotated matrix $\boldsymbol{N}$.

\item The same argument extends to the operator (spectral) norm:
$\min_{\operatorname{rank}(\boldsymbol{B})\leq k}\|\boldsymbol{A}-\boldsymbol{B}\| = \sigma_{k+1}$.
In that case, one uses the fact that the operator norm of $\boldsymbol{\Sigma}-\boldsymbol{N}$
is at least the absolute value of its largest diagonal entry that is not ``captured'' by $\boldsymbol{N}$.

\item The Eckart-Young theorem forms the theoretical foundation for PCA, low-rank matrix
completion, and many dimensionality reduction techniques.
\end{enumerate}

\newpage
%======================================================================
\section*{Question 4: LASSO Variant --- The Square-Root LASSO}
%======================================================================

\subsection*{4.1 Paper Details (2.5 points)}
\begin{itemize}
    \item \textbf{Title:} ``Square-Root Lasso: Pivotal Recovery of Sparse Signals via Conic Programming''
    \item \textbf{Authors:} Alexandre Belloni, Victor Chernozhukov, Lihong Wang
    \item \textbf{Venue:} \emph{Biometrika}, Volume 98, Issue 4
    \item \textbf{Year:} 2011
\end{itemize}

\subsection*{4.2 Mathematical Formulation (2.5 points)}
The standard LASSO solves
\[
\hat{\boldsymbol{\beta}}_{\text{LASSO}} = \arg\min_{\boldsymbol{\beta} \in \mathbb{R}^p}
\frac{1}{2n}\|\boldsymbol{y} - \boldsymbol{X}\boldsymbol{\beta}\|_2^2 + \lambda \|\boldsymbol{\beta}\|_1.
\]
The \textbf{Square-Root LASSO} replaces the squared loss with its square root:
\[
\boxed{\hat{\boldsymbol{\beta}}_{\text{SR}} = \arg\min_{\boldsymbol{\beta} \in \mathbb{R}^p}
\frac{1}{\sqrt{n}}\|\boldsymbol{y} - \boldsymbol{X}\boldsymbol{\beta}\|_2 + \lambda \|\boldsymbol{\beta}\|_1.}
\]

\paragraph{Notation and dimensions.}
\begin{itemize}
    \item $\boldsymbol{y} \in \mathbb{R}^n$: response vector ($n \times 1$).
    \item $\boldsymbol{X} \in \mathbb{R}^{n \times p}$: design matrix ($n$ observations, $p$ features).
    \item $\boldsymbol{\beta} \in \mathbb{R}^p$: coefficient vector ($p \times 1$).
    \item $\lambda > 0$: regularization parameter (scalar).
    \item $\|\cdot\|_2$: the Euclidean ($\ell_2$) norm.
    \item $\|\cdot\|_1$: the $\ell_1$ norm $\sum_{j=1}^p |\beta_j|$.
\end{itemize}

The model is $\boldsymbol{y} = \boldsymbol{X}\boldsymbol{\beta}^* + \boldsymbol{\varepsilon}$, where
$\boldsymbol{\varepsilon} \sim \mathcal{N}(\boldsymbol{0}, \sigma^2 \boldsymbol{I}_n)$ with \emph{unknown} $\sigma$.

\subsection*{4.3 Key Theorem (5 points)}

\begin{quote}
\textbf{Theorem (Belloni, Chernozhukov, Wang 2011).}
Suppose $\boldsymbol{y} = \boldsymbol{X}\boldsymbol{\beta}^* + \boldsymbol{\varepsilon}$ with $\boldsymbol{\varepsilon} \sim \mathcal{N}(\boldsymbol{0},\sigma^2\boldsymbol{I}_n)$.
Let $s = \|\boldsymbol{\beta}^*\|_0$ be the sparsity of the true coefficient vector. Choose the regularization parameter as
\[
\lambda = c \sqrt{\frac{\log p}{n}},
\]
where $c>0$ is a universal constant that does \emph{not} depend on $\sigma$.
Then, under a restricted eigenvalue condition on $\boldsymbol{X}$, the Square-Root LASSO estimator $\hat{\boldsymbol{\beta}}_{\text{SR}}$ satisfies
\[
\|\hat{\boldsymbol{\beta}}_{\text{SR}} - \boldsymbol{\beta}^*\|_2
\;\leq\; C\,\sigma\,\sqrt{\frac{s\,\log p}{n}},
\]
with high probability (at least $1 - O(1/p)$), where $C$ is a constant depending on the restricted eigenvalue of $\boldsymbol{X}$.
\end{quote}

The critical property is that $\lambda$ is \textbf{pivotal}: it does not depend on the unknown noise level $\sigma$. This is in contrast to the standard LASSO, where the optimal $\lambda$ scales as $\sigma\sqrt{\log p/n}$ and thus requires knowledge or estimation of $\sigma$.

\subsection*{4.4 Advantage over LASSO (10 points)}

The central advantage of the Square-Root LASSO over the standard LASSO is its
\textbf{pivotal} (self-normalizing) nature:

\begin{enumerate}
    \item \textbf{No need to know or estimate $\sigma$.}
    In the standard LASSO, the theoretically optimal regularization parameter is
    $\lambda_{\text{LASSO}} \propto \sigma\sqrt{\log p/n}$.
    Since the noise level $\sigma$ is almost never known in practice, one must either
    estimate it (which is itself a hard problem in the $p > n$ regime) or use
    computationally expensive cross-validation. The Square-Root LASSO removes this
    dependence entirely: $\lambda_{\text{SR}} \propto \sqrt{\log p/n}$, which depends only
    on the known quantities $n$ and $p$.

    \item \textbf{Equivalent near-oracle performance.}
    Despite the simpler tuning, the Square-Root LASSO achieves the same near-minimax
    optimal rate $O(\sigma\sqrt{s\log p/n})$ for estimation error as the optimally-tuned LASSO.
    There is no statistical price for the simpler calibration.

    \item \textbf{Practical robustness.}
    Because the estimator is pivotal, it is more robust to heteroscedasticity and
    model misspecification of the noise distribution. Even if $\sigma$ varies or is
    hard to estimate, the same $\lambda$ works.

    \item \textbf{Conic programming formulation.}
    The Square-Root LASSO can be reformulated as a second-order cone program (SOCP),
    which is a convex optimization problem solvable in polynomial time by standard
    interior-point methods. This makes it computationally tractable and allows the use
    of off-the-shelf solvers.
\end{enumerate}

In summary, the Square-Root LASSO retains all the theoretical guarantees of the LASSO
while eliminating the need for noise-level estimation, making it more practical and
easier to deploy in real-world high-dimensional settings.

\newpage
%======================================================================
\section*{Question 5: SVT for Low-Rank Matrix Completion}
%======================================================================

\subsection*{5.1 Problem Setup}

We solve the low-rank matrix completion problem via the Singular Value Thresholding
(SVT) algorithm.  Specifically, we use the Inexact Augmented Lagrange Multiplier (IALM)
variant, which converges significantly faster than vanilla SVT.

\paragraph{Matrix generation (Eckart-Young).}
A rank-$r$ matrix $\boldsymbol{X} \in \mathbb{R}^{n \times n}$ is created as
$\boldsymbol{X} = \boldsymbol{A}\boldsymbol{B}^T$ where
$\boldsymbol{A}, \boldsymbol{B} \in \mathbb{R}^{n \times r}$ have i.i.d.\
$\mathcal{N}(0,1)$ entries.

\paragraph{Observation model.}
We observe $m = f \cdot n^2$ entries uniformly at random and add i.i.d.\ Gaussian noise
with standard deviation $f_\sigma = 0.02 \times \text{mean}(|\text{observed entries}|)$.

\paragraph{Algorithm (IALM for nuclear norm minimization).}
We solve
\[
\min_{\boldsymbol{X}} \;\lambda\|\boldsymbol{X}\|_* \quad\text{s.t.}\quad
\mathcal{P}_\Omega(\boldsymbol{X}) = \boldsymbol{b},
\]
via the augmented Lagrangian iteration:
\begin{align}
\boldsymbol{T}^{(k)} &= \boldsymbol{X}^{(k)}\;\text{with}\;
T^{(k)}_{ij} \leftarrow b_{ij} + Y^{(k)}_{ij}/\mu_k \;\;\forall (i,j)\in\Omega,\\
\boldsymbol{X}^{(k+1)} &= \mathcal{D}_{\lambda/\mu_k}\!\left(\boldsymbol{T}^{(k)}\right)
\quad\text{(soft-threshold singular values)},\\
Y^{(k+1)}_{ij} &= Y^{(k)}_{ij} + \mu_k(b_{ij} - X^{(k+1)}_{ij}),\quad (i,j)\in\Omega,\\
\mu_{k+1} &= \min(\rho\,\mu_k,\; 10^7),
\end{align}
where $\mathcal{D}_\tau(\boldsymbol{M}) = \boldsymbol{U}\max(\boldsymbol{\Sigma}-\tau\boldsymbol{I},0)\boldsymbol{V}^T$
denotes the SVT operator, and $\rho = 1.15$ is the penalty growth factor.

\paragraph{Cross-validation.}
The regularization parameter $\lambda$ is selected by 2-fold cross-validation on the
observed entries, from the candidate set $\{0.01, 0.1, 1, 10, 50\}$.

\paragraph{RMSE metric.}
$\text{RMSE} = \|\hat{\boldsymbol{X}} - \boldsymbol{X}\|_F \;/\; \|\boldsymbol{X}\|_F$.

\subsection*{5.2 Results: $n = 1000$}

\begin{table}[h!]
\centering
\caption{RMSE $\|\hat{\boldsymbol{X}}-\boldsymbol{X}\|_F/\|\boldsymbol{X}\|_F$ for $n=1000$.}
\begin{tabular}{c|cccc}
\toprule
$r \;\backslash\; f$ & 0.05 & 0.15 & 0.30 & 0.40 \\
\midrule
3  & 0.2772 & 0.0434 & 0.0505 & 0.0538 \\
10 & 0.4320 & 0.0459 & 0.0532 & 0.0569 \\
40 & 0.9564 & 0.0538 & 0.0532 & 0.0563 \\
80 & 1.0008 & 0.6816 & 0.0522 & 0.0546 \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[h!]
\centering
\caption{Execution time (seconds) for $n=1000$.}
\begin{tabular}{c|cccc}
\toprule
$r \;\backslash\; f$ & 0.05 & 0.15 & 0.30 & 0.40 \\
\midrule
3  & 3.1 & 3.2 & 3.5 & 3.5 \\
10 & 4.5 & 5.0 & 5.8 & 4.6 \\
40 & 5.7 & 10.2 & 12.1 & 11.0 \\
80 & 3.2 & 8.9 & 10.9 & 10.8 \\
\bottomrule
\end{tabular}
\end{table}

\subsection*{5.3 Results: $n = 3000$}

\textbf{Note:} The original assignment leaves $r\in\{\}$ empty for $n=3000$.
We use $r \in \{3, 10, 40, 80\}$ (same as $n=1000$) as the most likely intended values.

\begin{table}[h!]
\centering
\caption{RMSE $\|\hat{\boldsymbol{X}}-\boldsymbol{X}\|_F/\|\boldsymbol{X}\|_F$ for $n=3000$.}
\begin{tabular}{c|cccc}
\toprule
$r \;\backslash\; f$ & 0.02 & 0.05 & 0.10 & 0.20 \\
\midrule
3  & 0.3657 & 0.1388 & 0.0584 & 0.0643 \\
10 & 0.4013 & 0.0883 & 0.0599 & 0.0675 \\
40 & 1.0117 & 0.1079 & 0.0614 & 0.0683 \\
80 & 1.0446 & 0.8918 & 0.0621 & 0.0690 \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[h!]
\centering
\caption{Execution time (seconds) for $n=3000$.}
\begin{tabular}{c|cccc}
\toprule
$r \;\backslash\; f$ & 0.02 & 0.05 & 0.10 & 0.20 \\
\midrule
3  & 27.8 & 30.1 & 31.4 & 38.1 \\
10 & 28.0 & 33.3 & 35.1 & 39.1 \\
40 & 33.5 & 47.4 & 46.8 & 52.6 \\
80 & 30.5 & 67.2 & 68.4 & 70.1 \\
\bottomrule
\end{tabular}
\end{table}

\subsection*{5.4 Discussion}

\begin{enumerate}
    \item \textbf{Effect of observation fraction $f$:}
    As $f$ increases (more entries observed), RMSE decreases significantly.
    For example, with $r=3$ at $n=1000$: RMSE drops from $0.277$ at $f=0.05$
    to $0.043$ at $f=0.15$.

    \item \textbf{Effect of rank $r$:}
    Higher rank matrices are harder to recover, especially when few entries are observed.
    At $f=0.05$, $r=80$ gives RMSE $\approx 1.0$ (essentially no recovery), while
    $r=3$ gives $0.277$. However, when $f$ is sufficiently large relative to $r$
    (the degrees of freedom of a rank-$r$ matrix is $r(2n-r)$), all ranks achieve
    similar low RMSE ($\sim 0.05$).

    \item \textbf{Information-theoretic limit:}
    A rank-$r$ matrix of size $n \times n$ has $r(2n-r)$ degrees of freedom.
    Recovery requires $m \gtrsim r(2n-r)$ observations. When $fn^2 < r(2n-r)$,
    recovery is fundamentally impossible, explaining the RMSE $\approx 1$ cases.

    \item \textbf{Execution time:}
    Time scales primarily with $n$ (due to SVD cost) and $k_{\text{svd}}$ (which
    tracks the rank). The IALM variant is efficient: $n=1000$ completes in $3$--$12$s
    and $n=3000$ in $28$--$70$s using randomised truncated SVD.
\end{enumerate}

\end{document}
